---
title: "Model cross-validation"
output: html_notebook
---

This is notebook runs the model cross-validation. In the cross-validation the data are divided in separate folds where one in turn is held out from the model estimation and it is used in testing how well the model predicts the results of unseen patients. Separate models are estimated for all the folds with different unseen patients as a test group. 

The data partitioning is done here by leaving out one blood test measurement from every tenth patient on weeks 4, 8 and 12. As the model uses one previous blood test measurement as a predictor, the week 0 is not considered in predictions. It is used as data for week 4 predictions.

Execution of this notebook takes about two days with a modern laptop and so this model generation is separated to this own notebook. All the intermediate results are stored in files and the execution continues from previous result if it is paused. The evaluation summaries are stored in evaluations/-folder from where they are included in the main notebook.

Same data files and model parameters are used here as in the model development 

```{r setup, include=FALSE}
library(knitr)
library(kableExtra) # for good looking tables

knitr::opts_chunk$set(echo = TRUE, fig.align="center")

# this allows using tikz rendering for plots with "dev=tikz"
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})

# Fix seed for random number generator for getting consistent results in kmeans etc.
fixed_seed <- 678

# Load common MEBN package
source("mebn/MEBN.r")
```

```{r data_loading, message=FALSE}

# Read the data description
datadesc <- read.csv(file="data/SYSDIMET_data_description.csv", header = TRUE, sep = ";")

# Read the actual data matching the description
sysdimet <- read.csv(file="data/SYSDIMET_diet.csv", sep=";", dec=",")

# Define how to iterate through the graph
assumedpredictors <- datadesc[datadesc$Order==100,]    
assumedtargets <- datadesc[datadesc$Order==200,] 
```

In the cross-validation two different model candidates are used: Gamma RHS AR(1) and Expfam RHS AR(1). The latter uses Normal distribution in modeling blood insulin. Same RHS parameters are used than before

```{r shrinkage_parameters, echo=TRUE, message=FALSE}
shrinkage_parameters <- within(list(),
{
    scale_icept  <- 1         # prior std for the intercept
    scale_global <- 0.02266   # scale for the half-t prior for tau: 
                              # ((p0=7) / (D=22-7)) * (sigma = 1 / sqrt(n=106*4))
    nu_global    <- 1         # degrees of freedom for the half-t priors for tau
    nu_local     <- 1         # degrees of freedom for the half-t priors for lambdas
    slab_scale   <- 1         # slab scale for the regularized horseshoe
    slab_df      <- 1         # slab degrees of freedom for the regularized horseshoe           
})
```

Following estimates cross-validation models for Gamma RHS AR(1). Change the 'cvmodelpath' variable to match a path in your environment. The path should have approximately 3 x 10 x 5 x 480 Mb = 72 empty space.

```{r 10-fold_holdout_gamma, echo=FALSE, eval=TRUE, message=FALSE}
library(igraph)

# Cross-validation models are stored in files at holdout-folders

for (week in c(4,8,12))
{
  for (holdout_suffix in as.character(0:9))
  {
    # Patients with 'holdout_suffix' as last characted in SUBJECT_ID are held out with their observation from 'week'
    holdout_index <- as.vector(as.numeric(endsWith(as.character(sysdimet$SUBJECT_ID), holdout_suffix) & sysdimet$WEEK == week))
    
    # Change this path to match your environment
    cvmodelpath <- "/media/jari/Jarin WD Passport"
    
    # This is relative path for model
    cvdir <- paste0(cvmodelpath, "/cvmodels/week", week,"/holdout_", holdout_suffix)
    
    if (!dir.exists(cvdir))
    {
      dir.create(cvdir)
    
      initial_graph <- mebn.new_graph_with_randomvariables(datadesc)
      sysdimet_gamma_ar1_rhs_cv <- mebn.bipartite_model(reaction_graph = initial_graph, 
                                       inputdata = sysdimet,
                                       targetdata = holdout_index,
                                       predictor_columns = assumedpredictors, 
                                       assumed_targets = assumedtargets, 
                                       group_column = "SUBJECT_ID",
                                       local_estimation = mebn.sampling,
                                       local_model_cache = cvdir, 
                                       stan_model_file = "mebn/BLMM_gamma_ar1_rhs_cv.stan",
                                       reg_params = shrinkage_parameters,
                                       normalize_values = TRUE)
      
      saveRDS(holdout_index, paste0(cvdir,"/holdout_index.rds"))
    }
  }
}
```

Following estimates only Normal distribution for blood insulin response. It is used as a replacement for Gamma model in Expfam RHS AR(1). This takes additional space of 3 x 10 x 1 x 480 Mb = 14.4 Gb.

```{r 10-fold_holdout_normal_fsins, echo=FALSE, eval=TRUE, message=FALSE}
library(igraph)
# Cross-validation models are stored in holdout-folders

fsins_only <- assumedtargets[assumedtargets$Name=="fsins",]

for (week in c(4,8,12))
{
  for (holdout_suffix in as.character(0:9))
  {
    # Patients with 'holdout_suffix' as last characted in SUBJECT_ID are held out with their observation from 'week'
    holdout_index <- as.vector(as.numeric(endsWith(as.character(sysdimet$SUBJECT_ID), holdout_suffix) & sysdimet$WEEK == week))
    
    # Change this path to match your environment
    cvmodelpath <- "/media/jari/Jarin WD Passport"
  
    cvdir <- paste0(cvmodelpath, "/cvmodels/normal_fsins/week", week,"/holdout_", holdout_suffix)
    
    if (dir.exists(cvdir))
    {
      dir.create(cvdir)
    
      initial_graph <- mebn.new_graph_with_randomvariables(datadesc)
      
      sysdimet_gamma_ar1_rhs_cv <- mebn.bipartite_model(reaction_graph = initial_graph, 
                                       inputdata = sysdimet,
                                       targetdata = holdout_index,
                                       predictor_columns = assumedpredictors, 
                                       assumed_targets = fsins_only, 
                                       group_column = "SUBJECT_ID",
                                       local_estimation = mebn.sampling,
                                       local_model_cache = cvdir, 
                                       stan_model_file = "mebn/BLMM_normal_ar1_rhs_cv.stan",
                                       reg_params = shrinkage_parameters,
                                       normalize_values = TRUE)
      
      saveRDS(holdout_index, paste0(cvdir,"/holdout_index.rds"))
    }
  }
}
```

Full Bayesian models are now estimated. They contain also predictions for unseen holdout patients. These predictions are now gathered in array with dimensions (blood test x week x patient). For each patient there is a predicted value of every modeled blood test response on weeks 4, 8, and 12. These results are stored in evaluations-folder.

Change cvmodelpath-variable to match your environment.

```{r collect_personal_predictions}
source("model_eval_functions.r")
source("mebn/MEBN.r")
library(dplyr)
library(stringr)

# This code collects personal predictions of blood tests in different study weeks from holdout models

# empty array with dimensions target,week,holdout
patients <- length(levels(sysdimet$SUBJECT_ID))

if (file.exists("evaluations/Gamma_pred_array.rds"))
{
  class.arr <- readRDS("evaluations/Gamma_pred_array.rds")
} else {
  class.arr <- array(-1, 
                   dim = c(nrow(assumedtargets), 3, patients),
                   dimnames = list(assumedtargets$Name, c(4,8,12), levels(sysdimet$SUBJECT_ID)))

  saveRDS(class.arr, "evaluations/Gamma_pred_array.rds")
}

for (week in c(4,8,12)) 
{
  for (holdout_suffix in as.character(0:9))
  {
      # Change this path to match your environment
      cvmodelpath <- "/media/jari/Jarin WD Passport"

      cvdir <- paste0(cvmodelpath, "/cvmodels/week", week, "/holdout_", holdout_suffix)
  
      for (targetname in assumedtargets$Name)
      {
        print(paste0(week, "/", holdout_suffix, "/", targetname))
        
        target_blmm <- mebn.get_localfit(targetname, cvdir)
  
        ms <- rstan::summary(target_blmm, pars=c("Y_pred"), probs=c(0.10, 0.90), na.rm = TRUE)
          
        # Insert predictions to array
        holdout_subjects <- unique(sysdimet[endsWith(as.character(sysdimet$SUBJECT_ID), holdout_suffix),]$SUBJECT_ID)
        
        # Extract prediction for all the holdout subjects for week and target
        pred_response <- as.vector(ms$summary[1:length(holdout_subjects),c(1)])
  
        # Insert to array
        class.arr[targetname,as.character(week),holdout_subjects] <- pred_response
        
        saveRDS(class.arr, "evaluations/Gamma_pred_array.rds")
      }
    }
}
```

Next the predictions are gathered for Expfam-model. Gamma-predictions are taken as a starting point and blood insulin predictions are replaced with Normal-distribution model.

Change cvmodelpath-variable to match your environment.

```{r collect_personal_predictions_ef}
# expfam
class.arr <- readRDS("evaluations/Gamma_pred_array.rds")

for (week in c(4,8,12)) 
{
  for (holdout_suffix in as.character(0:9))
  {
      # Change this path to match your environment
      cvmodelpath <- "/media/jari/Jarin WD Passport"
      
      cvdir <- paste0(cvmodelpath, "/cvmodels/normal_fsins/week", week, "/holdout_", holdout_suffix)
  
      targetname <- "fsins"
      print(paste0(week, "/", holdout_suffix, "/", targetname))
      
      target_blmm <- mebn.get_localfit(targetname, cvdir)

      ms <- rstan::summary(target_blmm, pars=c("Y_pred"), probs=c(0.10, 0.90), na.rm = TRUE)
        
      # Insert predictions to array
      holdout_subjects <- unique(sysdimet[endsWith(as.character(sysdimet$SUBJECT_ID), holdout_suffix),]$SUBJECT_ID)
      
      # Extract prediction for all the holdout subjects for week and target
      pred_response <- as.vector(ms$summary[1:length(holdout_subjects),c(1)])

      # Insert to array
      class.arr[targetname,as.character(week),holdout_subjects] <- pred_response
      
      saveRDS(class.arr, "evaluations/Expfam_pred_array.rds")
    }
}


```

Finally, the gathered predictions are compared with true values. Both NRMSE metrics and classification tests are computed for every patient and for every blood test. Means of these metrics are stored in blood test level and in overall model level. These summary matrices are stored in evaluations-folder and included in the main notebook. 

```{r NRMSE_and_accuracy_evaluation}
source("model_eval_functions.r")

gamma_pred.arr <- readRDS("evaluations/Gamma_pred_array.rds")
expfam_pred.arr <- readRDS("evaluations/Expfam_pred_array.rds")

patients <- length(levels(sysdimet$SUBJECT_ID))

gamma_nrmse.arr <- array(0, 
                         dim = c(nrow(assumedtargets), 4, patients),
                         dimnames = list(assumedtargets$Name, c('4', '8', '12', 'mean'), levels(sysdimet$SUBJECT_ID)))

expfam_nrmse.arr <- array(0, 
                         dim = c(nrow(assumedtargets), 4, patients),
                         dimnames = list(assumedtargets$Name, c('4', '8', '12', 'mean'), levels(sysdimet$SUBJECT_ID)))

gamma_class.mat <- matrix(0, nrow=nrow(assumedtargets),ncol=3)
expfam_class.mat <- matrix(0, nrow=nrow(assumedtargets),ncol=3)

number_of_preds <- 0
amount_subjects <- length(levels(sysdimet$SUBJECT_ID))

for (subject in levels(sysdimet$SUBJECT_ID))
{
    number_of_preds <- number_of_preds + 1

    # Get true values and different predictions
    gamma_response <- gamma_pred.arr[,,subject]
    expfam_response <- expfam_pred.arr[,,subject]
    true_response <- get_true_response_subj(subject)[,2:4]
    true_response0 <- get_true_response_subj(subject)
    
    # Classification accuracy
    
    true_delta <- make_delta(true_response0, true_response0)
    gamma_delta <- make_delta(true_response0, cbind(0, gamma_response))
    expfam_delta <- make_delta(true_response0, cbind(0, expfam_response))

    gamma_sign_matrix <- sign(gamma_delta) == sign(true_delta)
    expfam_sign_matrix <- sign(expfam_delta) == sign(true_delta)
    
    gamma_class.mat <- gamma_class.mat + gamma_sign_matrix[,2:4]*1
    expfam_class.mat <- expfam_class.mat + expfam_sign_matrix[,2:4]*1

    # NRMSE
    
    gamma_nrmse_rows <- mebn.NRMSE(gamma_response, true_response, mean(true_response))
    gamma_nrmse <- rowMeans(gamma_nrmse_rows)

    expfam_nrmse_rows <- mebn.NRMSE(expfam_response, true_response, mean(true_response))
    expfam_nrmse <- rowMeans(expfam_nrmse_rows)

    gamma_nrmse.arr[,,subject] <- cbind(gamma_nrmse_rows, gamma_nrmse)
    expfam_nrmse.arr[,,subject] <- cbind(expfam_nrmse_rows, expfam_nrmse)
}

# Classification accuracy

gamma_class_sum.mat <- gamma_class.mat/number_of_preds * 100
expfam_class_sum.mat <- expfam_class.mat/number_of_preds * 100

gamma_class_sum.mat <- cbind(gamma_class_sum.mat, rowMeans(gamma_class_sum.mat[,1:3]))
colnames(gamma_class_sum.mat) <- c('0 to 4 weeks', '4 to 8 weeks', '8 to 12 weeks', 'mean accuracy')
gamma_class_sum.mat <- round(gamma_class_sum.mat, 0)
saveRDS(gamma_class_sum.mat, "evaluations/GammaCV_accuracy_matrix.rds")

expfam_class_sum.mat <- cbind(expfam_class_sum.mat, rowMeans(expfam_class_sum.mat[,1:3]))
colnames(expfam_class_sum.mat) <- c('0 to 4 weeks', '4 to 8 weeks', '8 to 12 weeks', 'mean accuracy')
expfam_class_sum.mat <- round(expfam_class_sum.mat, 0)
saveRDS(expfam_class_sum.mat, "evaluations/ExpfamCV_accuracy_matrix.rds")

# NRMSE

gamma_eval <- c("Gamma AR(1) RHS CV", round(rowMeans(gamma_nrmse.arr[,4,]), 3), round(mean(rowMeans(gamma_nrmse.arr[,4,])), 3))
expfam_eval <- c("EXPFAM AR(1) RHS CV", round(rowMeans(expfam_nrmse.arr[,4,]), 3), round(mean(rowMeans(expfam_nrmse.arr[,4,])), 3))

nrmse.mat <- readRDS("evaluations/NRMSE-matrix2.rds")

nrmse.mat <- rbind(nrmse.mat, gamma_eval)
nrmse.mat <- rbind(nrmse.mat, expfam_eval)
rownames(nrmse.mat) <- rep("", nrow(nrmse.mat))

saveRDS(nrmse.mat, "evaluations/NRMSE-matrix3.rds")
```
